Live Functional Programming with Typed Holes

### 2018-04-10 calculus.tex
note about passive voice; 100 occurences of 'is' in this section\!
finished draft of sec3 and 4

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Matt's high-level Writing comments for Cyrus (for Section 3 mostly):

% Please, please please: Use line breaks.  Git does't work well without them.
% Change your editor or editor settings please.

% Cyrus, your writing is generally very good, but is often harder for
% me to digest than it should be, as a reader.  I think it'd be easier
% to read on a first pass if we followed my preferences below, which I
% humbly offer as gentle suggestions:

% I think that some these suggestions are especially important for reading
% really wide columns of text, as in this new PACMPL format. :)

% My nits boil down to these preferences:

% 0. NO PASSIVE VOICE --- EVER.  Use ``We'' in lots of places below instead of passive voice.

% 1. Shorter sentences.

% 2. Fewer parenthenticals: just use more sentences instead.

% 3. Shorter paragraphs, with more defined 'paragraph structure'. (E.g., topic sentence)

% 4. Shorter sentence subjects (with subject closer to verb).

% 5. Use macros for fig and section references.

% 6. If a sentence introduces a figure, that sentence should be the
% first sentence of a paragraph, and the figure reference should be
% the first word of that sentence, ideally.  As a reader, then I look
% at figures first, and find the paragraphs for them easily by
% scanning the text.

% 7. For each greek letter, prefix with an english word 'classifier',
% e.g., 'external expression~$e$', not just '$e$'.  Use way, way fewer
% commas please.

% 8. Use singlar over plural sentence subjects, whenever possible;
% e.g., an ``empty hole is written..'' not ``empty holes are
% written''; or ``an empty hole synthesizes'' not ``empty holes
% synthesize''

% 9. Fewer commas. Each 'extra' comma is a burden to the reader (it's
% asking the reader to allocate a stack frame in their brain!)

% *** I'm transforming the next paragraph below as a demonstration,
% creating smaller, more cohesive paragraphs with shorter sentences.
% See what you think.
%
% I'm not doing all of Section 3, tho, since
%
% (1) you may hate what I've done, and
%
% (2) I don't want to do a lot of work that will just be
% reverted/resented :)

% BTW: My version is _shorter_, not longer than the original :)
%
% Also, In my humble opinion, it is way, way easier to read. ;)

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{fig-bidirectional-typing}

We give an overview of the static semantics of the \HazelnutLive
external language.

\Figref{fig:bidirectional-typing} defines \emph{bidirectional} typing
%
as two mutually defined judgements: The type synthesis
judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
assumptions of the form $x : \htau$ in the usual
manner \cite{pfpl,tapl}.
%
The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
expression~$\hexp$ against a given type~$\htau$. Algorithmically, this
type is an output of type synthesis but an input of type analysis. At
the top level, we start with synthesis.

% Algorithmically, the type is an output of type synthesis but an input of type analysis.

The benefit of specifying the \HazelnutLive external language
bidirectionally is that the programmer does not need to annotate each
hole with a type
%
\cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}.
%
These typing rules closely follow the \Hazelnut type
system \cite{popl-paper}; we discuss minor differences below.

We supplement each expression hole from \Hazelnut with such a
hole \emph{name}, which are new to \HazelnutLive; an empty hole is
written $\hehole{u}$, where $u$ is a unique name for the hole.
%
Rule {SEHole}\todo{rule name macros}{} specifies that an empty hole synthesizes hole type, written $\tehole$.
%
If an empty hole appears where an expression of some other type is
expected, e.g. under an explicit ascription (governed by Rule {SAsc})
or in the argument position of a function application (governed by
Rule {SAp}, discussed below), we apply the \emph{subsumption rule},
Rule {ASubsume}, which specifies that if an expression $e$ synthesizes
type $\htau$, then it can be checked against any \emph{consistent}
type, $\htau'$.

\input{fig-type-consistency}

%%%%%%%%%%%%%%%%%%%% MATT stopped editing Cyrus's text here.

Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions. The hole type is consistent with every type, and so, by the subsumption rule, expression holes can appear where an expression of any type is expected. The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}. Note that type consistency is reflexive and symmetric but it is not transitive (unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).

@@

For the familiar forms of the lambda calculus, the rules again follow prior work. For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e. $b$ is the unit type). \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}. 
%
\rkc{perhaps too much info / unnecessary to say Hazelnut had numbers instead of unit?} 
%
Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

@@

Well-typed external expressions expand to well-typed internal expressions, $d$, for evaluation. The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}. Evaluation of internal expressions will be discussed in Sec.~\ref{sec:evaluation}.

Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. We say more about hole contexts below. The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$. The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
%
\rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
entirely standard, because of the hole context, so perhaps worth mentioning.}
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e. the following standard unicity property holds of the type assignment system: 
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces. For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

@@

The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} (so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation onto unannotated external lambdas based on the given type. The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.
%
\rkc{this list (holes, func app, and ascription) has three items, but there are
only two subsections. see comment at beginning of Sec 3.2.2.}

\subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively. 
%
\rkc{perhaps add ``empty and non-empty'' before hole closures.}

@@

Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.  The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. Initially, no evaluation has occurred, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

@@

\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
%
\rkc{consider renaming subsection to ``Function Application and Cast
Intersion''. and adding a transition sentence like ``The remaining development
of expansion deals with the delicate interplay between function application and
cast insertion.''}
%
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed. The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory. This leads us to the solution to this problem: cast insertion. 
%
\rkc{Instead of the last two sentences, maybe something like: ``By analogy
between hole types in our system and unknown types from gradual type theory, we
address the problem with the same mechanism used in gradual type theory: cast
insertion.''}

@@

Casts are inserted during the expansion of function applications and ascriptions. The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.
%
\rkc{move this Note sentence to the same place where half-annotated lambdas were
noted. This note breaks the flow here.}

@@

The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}). The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}

@@

Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. The only cases of special interest are when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write $[d/x]\sigma$ to perform substitution on each expression in the hole environment, $\sigma$. For example, $\stepsToD{}{\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}{\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g. because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?}. Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.

@@

The last possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$. In this case, no further progress can be made, so the function application as a whole is also deemed indeterminate by Rule {IAp} in Fig.~\ref{fig:isFinal}. Note that we are careful to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.\todo{prove this as stated from Ian's disjointness proofs}\todo{do we need a typing premise?}
\begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
\end{thm}

@@

\subsubsection{Type Safety} afdasfdas dfasdfs
\begin{thm}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$.
\end{thm}

\begin{thm}[Progress]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
  (i) $\stepsToD{}{\dexp}{\dexp'}$ or
  (ii) $\isIndet{\dexp}$ or
  (iii) $\isBoxedValue{\dexp}$.
\end{thm}

\begin{thm}[Canonical Boxed Forms]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$ and $\isBoxedValue{\dexp}$
  then
  \begin{enumerate}[label=(\alph*)]
    \item If $\htau=b$ then $\dexp=c$.
    \item If $\htau=\tarr{\htau_1}{\htau_2}$ then either
      \begin{enumerate}
        \item[i.]
          $\dexp=\halam{x}{\htau_1}{\dexp'}$
          where $\hasType{\hDelta}{x : \htau_1}{\dexp'}{\htau_2}$, or
        \item[ii.]
          $\dexp=\dcasttwo{\dexp'}{\tarr{\htau_1'}{\htau_2'}}{\tarr{\htau_1}{\htau_2}}$
          where $\tarr{\htau_1'}{\htau_2'}\neq\tarr{\htau_1}{\htau_2}$
          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\tarr{\htau_1'}{\htau_2'}}$.
      \end{enumerate}
    \item If $\htau=\tehole$
          then $\dexp=\dcasttwo{\dexp'}{\htau'}{\tehole}$
          where $\isGround{\htau'}$
          and $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau'}$.
  \end{enumerate}
\end{thm}

>>>>

\input{fig-bidirectional-typing}
\input{fig-type-consistency}

Let us start with an overview of the static semantics of the \HazelnutLive external language. Types, $\htau$, classify external expressions, $e$, according to the type system in Fig. \ref{fig:bidirectional-typing}, which closely follows the \Hazelnut type system \cite{popl-paper} (the minor differences are discussed as they come up below). The type system is specified in the \emph{bidirectional} style \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da} around two mutually defined judgements. The type synthesis judgement, $\hsyn{\hGamma}{\hexp}{\htau}$, synthesizes a type $\htau$ for $\hexp$ under typing context $\hGamma$, which tracks typing assumptions of the form $x : \htau$ in the usual manner \cite{pfpl,tapl}. The type analysis judgement, $\hana{\hGamma}{\hexp}{\htau}$, instead checks $\hexp$ against a given $\htau$. Algorithmically, the type is an output of type synthesis but an input of type analysis. At the top level, we start with synthesis. % Algorithmically, the type is an output of type synthesis but an input of type analysis.

The benefit of specifying the \HazelnutLive external language bidirectionally is that the programmer does not need to annotate each hole with a type. Empty holes are written $\hehole{u}$, where $u$ is a hole name assumed unique (hole notation is taken from \Hazelnut; hole names are new to \HazelnutLive). Rule {SEHole}\todo{rule name macros}{} specifies that empty holes synthesize hole type, written $\tehole$. If an empty hole appears where an expression of some other type is expected, e.g. under an explicit ascription (governed by Rule {SAsc}) or in the argument position of a function application (governed by Rule {SAp}, discussed below), we apply the \emph{subsumption rule}, Rule {ASubsume}, which specifies that if an expression $e$ synthesizes type $\htau$, then it can be checked against any \emph{consistent} type, $\htau'$. Fig.~\ref{fig:tconsistent} specifies the type consistency relation, written $\tconsistent{\htau}{\htau'}$, which specifies that two types are consistent if they differ only up to type holes in corresponding positions. The hole type is consistent with every type, and so, by the subsumption rule, expression holes can appear where an expression of any type is expected. The type consistency relation here coincides with the type consistency relation from gradual type theory by identifying the hole type with the unknown type~\cite{Siek06a}. Note that type consistency is reflexive and symmetric but it is not transitive (unlike subtyping, which is anti-symmetric and transitive; subtyping can be integrated into a gradual type system following \citet{Siek:2007qy}).

@@

For the familiar forms of the lambda calculus, the rules again follow prior work. For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e. $b$ is the unit type). \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.

Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

@@

Well-typed external expressions expand to well-typed internal expressions, $d$, for evaluation. The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}. We will return to evaluation in Sec.~\ref{sec:evaluation}.

Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. We say more about hole contexts below. The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$. The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e. the following standard unicity property holds of the type assignment system: 
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces. For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

@@

The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} (so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation onto unannotated external lambdas based on the given type. The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.

@@

Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$. The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. Initially, no evaluation has occured, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

@@

\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed. The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory. This leads us to the solution to this problem: cast insertion.

@@

Casts are inserted during the expansion of function applications and ascriptions. The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.

@@

The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}. The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}

@@

Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. The only cases of special interest are when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write $[d/x]\sigma$ to perform substitution on each expression in the hole environment, $\sigma$. For example, $\stepsToD{}{\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}{\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g. because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?} Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.

@@

The other possibility is that the expression in function position is indeterminate and arrow cast conversion is not applicable, e.g. $\hap{(\dehole{u}{\sigma}{})}{c}$. In this case, no further progress can be made, so the function application as a whole is also deemed indeterminate by Rule {IAp} in Fig.~\ref{fig:isFinal}. Note that we are careful to maintain the following property, which establishes that expressions that step are not final, and that final expressions do not step.\todo{prove this as stated from Ian's disjointness proofs}\todo{do we need a typing premise?}
\begin{thm}[Finality] There does not exist $d$ such that both $\isFinal{d}$ and $\stepsToD{}{d}{d'}$ for some $d'$.
\end{thm}

@@

In all other cases, casts are either boxed values or indeterminate according to the remaining rules in Fig.~\ref{fig:isFinal}. Note in particular Rule {ICastHoleGround}, which handles casts from hole to ground type that are not of the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$. In the cast calculus, there are no such irreducible terms because the only canonical form of type $\tehole$ is $\dcasttwo{\dexp}{\htau_1}{\tehole}$.

\subsubsection{Type Safety} The purpose of establishing type safety is to ensure that the static and dynamic semantics of a language cohere. We follow the approach developed by \citet{wright94:_type_soundness}, now standard \cite{pfpl}, which distinguishes two type safety properties, preservation and progress. In order to allow for the evaluation of incomplete programs, we establish these properties for terms typed under arbitrary hole context $\Delta$. We can assume $\hGamma$ is empty (in order to run programs with free variables, the system can treat them as empty holes with a corresponding name).

The preservation theorem establishes that type assignment is preserved by transitions, i.e. that the type of an expression accurately predicts the type of the result of evaluating that expression.

\begin{thm}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$.
\end{thm}
\noindent
The proof relies on an analagous preservation lemma for instruction transitions and a standard substitution lemma. Hole closures can disappear during evaluation, so we rely on weakening of $\Delta$.

The progress theorem establishes that the dynamic semantics accounts for every well-typed term, i.e. that we have not forgotten some necessary rules or premises.
\begin{thm}[Progress]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
  (a) $\stepsToD{}{\dexp}{\dexp'}$ or
  (b) $\isBoxedValue{\dexp}$ or 
  (c) $\isIndet{\dexp}$.
\end{thm}
\noindent
Explicitly accounting for indeterminate forms, which are rooted at either a hole closure or a failed cast, is the key to being able to establish the progress theorem under a non-empty hole context. The proof relies on canonical forms lemmas stated in Appendix~\ref{sec:additional-defns}.

\subsubsection{Complete Programs} Although our focus in this paper is on running incomplete programs, this machinery does not interfere with reasoning about the behavior of complete programs, i.e. those with no holes. Appendix~\ref{sec:additional-defns} defines complete programs with the judgements $\isComplete{e}$ and $\isComplete{d}$. Failed casts cannot appear in complete internal expressions. The following theorem establishes that expansion preserves program completeness.
\begin{thm}[Complete Expansion] ~
  \begin{enumerate}[nolistsep]
    \item
      If $\isComplete{\hexp}$
      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\dexp}$.
    \item
      If $\isComplete{\hexp}$
      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\isComplete{\dexp}$.
  \end{enumerate}
\end{thm}

The following preservation theorem establishes that stepping preserves program completeness.\todo{can we do complete preservation without the typing assumption?}
\begin{thm}[Complete Preservation]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
  and $\isComplete{\dexp}$
  and $\stepsToD{}{\dexp}{\dexp'}$
  then $\hasType{\hDelta}{\emptyset}{\dexp'}{\htau}$
  and $\isComplete{\dexp'}$.
\end{thm}

The following progress theorem establishes that only classic values result from evaluating a complete program. There are no boxed values or indeterminate forms. 
\begin{thm}[Complete Progress]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
  and $\isComplete{\dexp}$
  then either $\stepsToD{}{\dexp}{\dexp'}$
  or $\isValue{\dexp}$.
\end{thm}
